---
title: "Chronic Disease Surveillance"
author: "Jungmin Chang"
date: "January 19, 2021"
output: html_document
---

```{r setup, include=F, message=F}
options(gsubfn.engine = "R")
library("knitr")
library("sqldf")
library("dplyr")
library("VennDiagram")
library("ggplot2")

# Set your working directory as needed.
# opts_knit$set(root.dir = "", fig.align = "center")
```
```{r data, echo=F}
hospital_discharges = read.csv('data/hospital_discharges.csv') # fee for service - claims for physicians who saw include diagnostic using icd-9-cm related to diabetes (only single code for 1 and 2)--simplied version of dischanrge - record o fadmissino with diagnosistic and reason for admission - ppl can appear multiple time or never
physician_services = read.csv('data/physician_services.csv') #
sampled_patients = read.csv('data/sampled_patients.csv') #-- denominator who is insured - each patients appear once!

hospital_discharges$admit = as.Date(hospital_discharges$admit)
hospital_discharges$discharge = as.Date(hospital_discharges$discharge)
physician_services$date = as.Date(physician_services$date)
sampled_patients$dob = as.Date(paste(as.character(sampled_patients$dob), "-01", sep = ""), format = "%Y-%m-%d")

str(hospital_discharges)
str(physician_services)
str(sampled_patients)
```

## Diabetes Mellitus Case Definition

Popular algorithm (Hux et al, 2002):

* Two physician diabetes diagnosis codes separated by 730 days or less OR 
* One hospitalization diabetes diagnosis code.

| Measure      | Performance |
|:-------------|:------------|
| Sensitivity: | 86%         | 
| Specificity: | 97%         |

Codes:

* ICD-9:250.*
* ICD-10: E10\*-E14\*

## 1. Identify people with at least one hospital admission for diabetes.

Steps

* We take the date of first admission only (i.e., "min(admit)"), so there will be only one row for each person
* When an aggregtion operator is used in the SELECT clause, there must be a GROUP BY 
* Ranges of ICD9 and ICD10 codes are given to account for the change in coding

```{r q1_1}
hospital_diag = sqldf(
  "SELECT anon_id, min(admit) AS diab_date 
   FROM hospital_discharges 
   WHERE (icd_type='ICD-9' AND icd LIKE '250%')
      OR (icd_type='ICD-10' AND (icd LIKE 'E10%' OR icd LIKE 'E11%' OR
                                 icd LIKE 'E12%' OR icd LIKE 'E13%' OR
                                 icd LIKE 'E14%'))
    GROUP BY anon_id")
```

* Identify all hospitalization events for diabetes
* We take all admissions, so there will be multiple rows for a person if they had multiple admissions

```{r q1_2}
hospital_diag_events = sqldf(
  "SELECT anon_id, admit AS diab_date 
   FROM hospital_discharges 
   WHERE (icd_type='ICD-9'  AND icd LIKE '250%')
      OR (icd_type='ICD-10' AND (icd LIKE 'E10%' OR icd LIKE 'E11%' OR
                                 icd LIKE 'E12%' OR icd LIKE 'E13%' OR
                                 icd LIKE 'E14%'))")
```


**Question A -  What proportion of all subjects had at least one admission for diabetes (R)?**

```{r q1a}
# Place any additional code here
nrow(hospital_diag) / nrow(sampled_patients)
```
2.64% of all subjects had at least one admission for diabetes

**Question B - Plot the frequency distribution of the number of hospitalizations per person for diabetes (SQL + R).**
__Hint: This is easy if you perform another SQL query very similar to the one for hospital_diag...__
```{r q1b}
# Place any additional code here
hospital_diag_distribution = sqldf(
  "SELECT anon_id, count(admit) AS ttl_hosp 
   FROM hospital_discharges 
   WHERE (icd_type='ICD-9' AND icd LIKE '250%')
      OR (icd_type='ICD-10' AND (icd LIKE 'E10%' OR icd LIKE 'E11%' OR
                                 icd LIKE 'E12%' OR icd LIKE 'E13%' OR
                                 icd LIKE 'E14%'))
    GROUP BY anon_id")

hist(hospital_diag_distribution$ttl_hosp, xlab = "ID", ylab ="Total number of hospitalization per person for disabetes", main = "The distribution of hospitalization for diabetes", cex.main = 0.8, cex.lab = 0.8)
```

**Question C - What proportion of all hospital admissions were for diabetes (R)?**
```{r q1c}
# Place any additional code here 
nrow(hospital_diag_events) / nrow(hospital_discharges)
```
2.62% of all hospital admissions were for diabetes

**Question D - When was ICD-10 first used for coding hospital discharges (SQL)?**
```{r q1d}
# Place any additional code here -- using sql 
first_icd10 = sqldf(
  "SELECT admit, discharge, icd, icd_type
   FROM hospital_discharges 
   WHERE icd_type = 'ICD-10'
   ORDER BY discharge")

min(hospital_discharges$discharge[hospital_discharges$icd_type== "ICD-10"])
```
ICD-10 code was used for coding hospital discharges as of March 29th, 2006

**Question E - Compare the proportion of hospital admissions that were for diabetes before and after the switch to ICD-10 (SQL + R).**

__Rates Before...__
```{r q1e1}
hospital_diag_preICD10 = sqldf(
  "SELECT anon_id, discharge
   FROM hospital_discharges
   WHERE icd_type='ICD-9' 
     AND icd LIKE '250%'")

hospital_discharge_count_preICD10 = sqldf(
  "SELECT COUNT(*)
   FROM hospital_discharges
   WHERE icd_type='ICD-9'")[1,1]  

nrow(hospital_diag_preICD10) / hospital_discharge_count_preICD10
```


__Rates after...__

```{r q1e2}
# Place any additional code here
hospital_diag_postICD10 = sqldf(
  "SELECT anon_id, discharge
   FROM hospital_discharges
   WHERE icd_type='ICD-10' 
     AND icd LIKE 'E10%' 
     OR icd LIKE 'E11%' 
     OR icd LIKE 'E12%' 
     OR icd LIKE 'E13%' 
     OR icd LIKE 'E14%'")

hospital_discharge_count_postICD10 = sqldf(
  "SELECT COUNT(*)
   FROM hospital_discharges
   WHERE icd_type='ICD-10'")[1,1]  

nrow(hospital_diag_postICD10) / hospital_discharge_count_postICD10
```
The proportion of hospital admissions that were for diabetes was 2.68% before the switch to ICD-10 code. The proportion became smaller after switching to ICD-10 code, which was approximately 2.13%

## 2 - Identify physician billing events for diabetes

```{r q2}
phys_diab = sqldf(
  "SELECT anon_id, date 
   FROM physician_services 
   WHERE icd LIKE '250%'")

# Physician consultation rate
phys_diab_unique = sqldf(
  "SELECT DISTINCT anon_id
   FROM physician_services
   WHERE icd LIKE '250%'")
```

**Question A - What proportion of all subjects had at least one physician visit for diabetes (R)?**
```{r q2e1}
# Place any additional code here
nrow(phys_diab_unique) / nrow(sampled_patients)
```
Approximately 5.77% of all subjects had at least one physician visit for diabetes

**Question B - Plot the frequency distribution of physician visits per person for diabetes (SQL + R).**
```{r q2e2}
# Place any additional code here
phys_diab_distribution = sqldf(
"SELECT anon_id, count(anon_id) as ttl_visit
   FROM physician_services 
   WHERE icd LIKE '250%'
   GROUP BY anon_id"
)

hist(phys_diab_distribution$ttl_visit, xlab = "ID", ylab ="Total number of physician visits per person for disabetes", main = "The distribution of physician visits for diabetes", cex.main = 0.8, cex.lab = 0.8)
```

**Question C - What proportion of all physician visits were for diabetes (R)?**
```{r q2e3}
# Place any additional code here
phys_all = sqldf(
  "SELECT count(anon_id) as all_physician_visits
   FROM physician_services")

nrow(phys_diab) / 336804 #2.01%
```
Approximately 2.01% of all physician visits were for diabetes

## 3 - Identify people with two or more billings for diabetes within 730 days

```{r q3}
phys_diag = sqldf(
  "SELECT x.anon_id, (x.date - y.date) as interval, min(x.date) as diab_date
   FROM phys_diab x 
   JOIN phys_diab y ON x.anon_id=y.anon_id 
                   AND x.date > y.date 
                   AND (x.date - y.date <=730) 
   GROUP BY x.anon_id")
```

**Question A - What proportion of subjects had two visits for diabetes within 730 days?**
```{r q3e1}
# Place any additional code here
nrow(phys_diag) / nrow(sampled_patients)
```
Approximately 4.45% of subjects had two visits for diabetes within 730 days

**Question B - Plot the proportion of subjects that would be identified in 3a as a function of the cutoff ranging from 0 to 730 days. Do you think 730 days is a reasonable cut-off? Explain.**
```{r q3e2}
# Place any additional code here
phys_diag1 = sqldf(
  "SELECT x.anon_id, (x.date - y.date) as interval, min(x.date) as diab_date
   FROM phys_diab x 
   JOIN phys_diab y ON x.anon_id=y.anon_id 
                   AND x.date > y.date
   GROUP BY x.anon_id
   HAVING interval <= 730")
phys_diag2 = sqldf(
  "SELECT x.anon_id, (x.date - y.date) as interval, min(x.date) as diab_date
   FROM phys_diab x 
   JOIN phys_diab y ON x.anon_id=y.anon_id 
                   AND x.date > y.date
   GROUP BY x.anon_id")




hist(phys_diag1$interval, freq = FALSE, xlab = "Interval", ylab ="Density", main = "The proportion of patients with \n two physician visits within 730 days", cex.main = 0.8, cex.lab = 0.8, cex.axis = 0.8, breaks = 15)
abline(v = 730, col = "red")

hist(phys_diag2$interval, freq = FALSE, xlab = "Interval", ylab ="Density", main = "Probability Density Function", cex.main = 0.8, cex.lab = 0.8, cex.axis = 0.8, breaks = 50)
abline(v = 730, col = "red")

```
I think 730 days is a reasonable cutoff because most of physician visits (more than two) occurred within 2-year period. However, 730 days cutoff might have ignored some cases which occurred after 2 years. 


## 4 - Join cases detected through physician billing with those detected from hospital discharges.

```{r q4}
both_diag = sqldf(
  "SELECT anon_id, diab_date FROM phys_diag 
   UNION
   SELECT anon_id, diab_date FROM hospital_diag")
                   
diab_dates = sqldf(
  "SELECT anon_id, min(diab_date) as diab_date 
   FROM both_diag 
   GROUP BY anon_id")

```

**Question A - What proportion of subjects met the case definition for diabetes of either one hospital admission or two physician visits within 730 days (R)?**
```{r q4e1}
# Place any additional code here
nrow(diab_dates) / nrow(sampled_patients)
```
Approximately 5.37% of subjects met the case definition for diabetes of either one hospital admission or two physician visits within 730 days

**Question B - Create a Venn diagram (the areas does not need to perfectly reflect the proportions) to illustrate the following: total number of patients, number that are cases due to physician billing, hospital discharges, or both (R).**
```{r q4e2}

# Place any additional code here
nrow(both_diag) #--709
nrow(diab_dates) #--537
nrow(phys_diag)
nrow(hospital_diag)
draw.triple.venn(area1 = 445,
                 area2 = 264,
                 area3 = 10000,
                 n12 = 172, 
                 n23 = 264,
                 n13 = 445,
                 n123 = 172,
                category = c("physcian billing", "hospital discharges", "total number of patients"),
                cat.pos = c(-40, 40, 180), 
                cat.dist = c(0.05, 0.05, 0.025),
                 cat.cex = rep(0.8,3),
                   cat.fontface = rep("bold",3),
                   fill = c("purple", "green", "honeydew2" ),
                   col =  c("purple", "green", "honeydew2"))
nrow(sampled_patients)
```

**Question C - Calculate the proportion of subjects that meet the cased definition stratified by age (44 and under, 45 to 64, 65 and over) and sex (male, female) stratified rates. (Hint - you will need to join to the patients table) (SQL + R).**
```{r q4e3}
# Place any additional code here
strata <- sqldf(
   "SELECT a.diab_date, a.anon_id, b.dob, b.sex
   FROM diab_dates AS a
   JOIN sampled_patients AS b
   ON a.anon_id = b.anon_id"
)

lubridate::as_date(0) # origin 1970-01-01
strata$diab_date <- format(as.Date(strata$diab_date, origin = "1970-01-01" ), format = "%Y")
strata$dob <- format(as.Date(strata$dob), format = "%Y")
strata$age_discrete <- as.numeric(strata$diab_date) - as.numeric(strata$dob)

strata$age_category <- ifelse(strata$age_discrete <=44, 0,
                              ifelse(strata$age_discrete > 44 & strata$age_discrete <=64,1, 
                                     ifelse(strata$age_discrete >64, 2,3)))

strata_sex <- strata %>%
   group_by(sex) %>%
   summarize(freq = round(n()/nrow(strata),2))

strata_age <- strata %>%
   group_by(age_category) %>%
   summarize(freq = round(n()/nrow(strata),2))


pie(strata_sex$freq, labels = strata_sex$freq, col = c("purple", "green"),lty = "blank", main = "Sex distribution of subjects that meet the case definition", cex.main = 0.8, cex = 0.8)
legend("topright", c("female", "male"), fill = c("purple", "green"), cex = 0.8)

pie(strata_age$freq, labels = strata_age$freq, col = c("purple", "green", "deeppink"), lty = "blank", main = "Age distribution of subjects that meet the case definition", cex.main = 0.8, cex = 0.8)
legend("topright", c("44 and under", "45 to 64", "65 and over"), fill = c("purple", "green", "deeppink"), cex = 0.8)

```
In order to stratify the proportion of subjects that met the case definition by age and sex, we joined two data sets, *diab_date* and *sampled_patient*, including sex and date of birth columns. Using this merged data, we categorized patients into three different age groups (age at hospitalization/physician visits).

The proportion of female and male patients was more or less the same. The proportion of male patients was slightly higher (52%) than female patients (48%).

Patients aged 65 or more accounted for the highest proportion (47%), followed by patients aged 45 to 64. Only 10% of patients who met the case definition were aged 44 and under.
